BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-30 ms.tex

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figures/domain_sampling.pdf}
    \caption{A $32\times 64$ grid in $(\alpha, \chi)$ mapped to the $(\hat{a}, \hat{\Omega})$ domain. Each dot represents a sampled point in the parameter space. These points are equally-spaced in $\alpha$ and in $\chi$, but cluster around $\hat{\Omega} = \hat{\Omega}_\mathrm{ISCO}$ and $\hat{a} = \hat{a}_\mathrm{max}$. The shading of each point reflects the relative density of points in the $(\hat{a}, \hat{\Omega})$ domain, while the dashed lines represent the domain boundaries.}
    \label{fig:domain}
\end{figure}

@@

Similar to the trajectories, we presample the complex waveform amplitudes $B_{\ell m}$ on a fixed $65 \times 65$ grid in $(\alpha, \chi)$. To optimize the accuracy of our amplitude interpolation, we decompose $B_{\ell m}$ into its phase $\psi_{\ell m}$ and the log of its real amplitude $\ln A_{\ell m}$. We then interpolate these quantities independently. We generate mode data for $\ell \leq 20$ and $0 < m \leq \ell$. Even with the reduced sampling, the interpolation errors introduce estimated fractional errors $< 3\times 10^{-5}$ in $A_{\ell m}$ and absolute errors $<2 \times 10^{-5}$ for $\psi_{\ell m}$, as discussed in Appendix \ref{app:tests}.

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency, but for $\hat{a} \gtrsim 0.995$ the amplitudes can peak well before reaching the ISCO. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO, which is highlighted in the middle panel of \ref{fig:amplitudeVariation} for all $m =2$ modes. This behavior is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. Furthermore, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 15$ modes for a given orbital frequency. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot the $A_{\ell, m = \ell}$ amplitudes as functions of $r_0$ for different values of $\hat{a}$. We see that $|A_{15,15}/A_{2,2}|^2 \lesssim 5\times 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 15$ mode at a fixed frequency. For most year-long EMRI signals, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 15$ mode is $\lesssim 10^{-6}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell = m = 20$, even for rapidly-spinning black holes.

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/amp_combined.pdf}
    \caption{The top panel plots the waveform amplitude $A_{\ell=2, m=2}$ as a function of the radius of the orbit $r_0/M$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The bottom panel plots the $A_{\ell, m=\ell}$ modes as a function of $r_0/M$ for $\hat{a} = 0.9999$ and all $2 \leq \ell \leq 15$. The magnitude of the $\ell = m$ modes is ordered inversely with $\ell$ in the bottom panel. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 15$ being the mode with the smallest magnitude. The bottom (yellow) line in the top panel is identical to the top (yellow) line in the bottom panel.}
    \label{fig:amplitudeVariation}
\end{figure}

@@

We presented the theoretical and numerical methods behind \texttt{bhpwave}: a new Python-based adiabatic gravitational waveform generator for binary sources composed of a compact object undergoing a quasi-circular inspiral into a rotating massive black hole. To build this waveform model, we precomputed mass-independent trajectories for systems with initial separations $r_0/M \lesssim 60$ and Kerr spins $|a| \leq 0.9999$. Furthermore, we precomputed all potential waveform harmonic amplitudes for $\ell \leq 20$. By implementing the $E(3)$ boundary conditions in our numerical spline algorithm, we improved the precision of our interpolated flux data and trajectories by as much as three orders of magnitude over traditional spline methods. We compute waveforms in both the time and frequency domains, and we find good agreement between these models. Furthermore, we compared our model against the gold-standard FEW waveform generator for $\hat{a} = e_0 = 0$ and achieved mismatches $\mathcal{M} \sim 10^{-5}$. Using Fisher matrix calculations, we also assessed the magnitudes of any systematic biases introduced by numerical error in our model. We found that biases due to interpolation error are well below the thresholds required for LISA data analysis. Additionally, we demonstrated that waveform dephasing does not provide a complete picture of modeling error. In particular, for systems with retrograde orbits and slowly-rotating MBHs, we found that waveform models could have phase errors of up to 10 radians, and yet the biases introduced by these errors would not be measurable by LISA. Thus, the claim that waveform models require subradian phase accuracy for LISA is a useful guide for model fidelity, but more sophisticated analyses are required to understand the true impact of modeling errors on LISA science, as most recently demonstrated in \cite{BurkETC23}.

A notable limitation of \texttt{bhpwave} is that it currently neglects the effects of eccentricity and precession.\footnote{In a frame where the massive black hole's angular momentum is held fixed, precession is instead described in terms of the inclination of the small body relative to the equatorial plane of the more massive black hole.} Nonetheless, while most observed EMRIs are expected to be highly eccentric \cite{GairETC04},  there are possible formation channels driven by accretion flow that circularize EMRI dynamics and align the orbital angular momentum with the massive black hole's spin \cite{PanLyuYang21}. Thus, \texttt{bhpwave} is applicable to these so-called ``wet-formation" EMRIs.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{figures/domain_sampling.pdf}
    \caption{A $32\times 64$ grid in $(\alpha, \chi)$ mapped to the $(\hat{a}, \hat{\Omega})$ domain.  Each dot represents a sampled point in the parameter space. These points are equally-spaced in $\alpha$ and in $\chi$, but cluster around $\hat{\Omega} = \hat{\Omega}_\mathrm{ISCO}$ and $\hat{a} = \hat{a}_\mathrm{max}$. The shading of each point reflects the relative density of points in the $(\hat{a}, \hat{\Omega})$ domain, while the dashed lines represent the domain boundaries. Note that this is downsampled version of the numerical grid used for interpolating flux and trajectory information within \texttt{bhpwave}.}
    \label{fig:domain}
\end{figure}

@@

Similar to the trajectories, we presample the complex waveform amplitudes $B_{\ell m}$ on a fixed $65 \times 65$ grid in $(\alpha, \chi)$. To optimize the accuracy of our amplitude interpolation, we decompose $B_{\ell m}$ into its phase $\psi_{\ell m}$ and the log of its real amplitude $\ln A_{\ell m}$. We then interpolate these quantities independently. We generate mode data for $\ell \leq 15$ and $0 < m \leq \ell$. Even with the reduced sampling, the interpolation errors introduce estimated fractional errors $< 3\times 10^{-5}$ in $A_{\ell m}$ and absolute errors $<2 \times 10^{-5}$ for $\psi_{\ell m}$, as discussed in Appendix \ref{app:tests}.

The amplitudes display a number of interesting properties, particularly as we move into the near-horizon extremal Kerr (NHEK) regime, i.e., $1-\hat{a}^2 \ll 1$ and $(r-r_+)/r_+ \ll 1$. For $\hat{a} \lesssim 0.995$ the amplitudes peak at the ISCO and decay with the orbital frequency. For $\hat{a} \gtrsim 0.995$ the amplitudes can instead peak well before reaching the ISCO and then decay as the orbital frequency increases. This behavior has been reported in previous investigations of extremal Kerr black holes \cite{GralPorfWarb15, GralHughWarb16}. For the dominant $\ell = m = 2$ mode, the waveform amplitudes reach their maximum magnitudes near the stationary surface of the outer ergosphere $r_E^+ = 2M$, as demonstrated in the top panel of Figure \ref{fig:amplitudeVariation}. The modes do not have the same turnover behavior for $\ell \neq m$, but instead plateau around $r_E^+$ before rising again as they approach the ISCO, which is highlighted in the middle panel of Figure \ref{fig:amplitudeVariation} for all $m =2$ modes. This behavior is due to our reprojection of the $(j, m)$-modes of a spin-weighted spheroidal harmonic basis to the $(\ell, m)$-modes of a spherical basis, given in \eqref{eqn:XfromR}. As demonstrated in \cite{GralPorfWarb15}, larger spheroidal $j$ modes (for a fixed $m$) will decay more rapidly as one approaches the ISCO in the NHEK regime. Therefore, in the spherical $(\ell, m)$-basis, the near-ISCO behavior is dominated by the contribution from the modes around $j=m$, even when $\ell \gg m$. As a result, for $\ell > m$ the spherical-spheroidal mixing with many $j\sim m$ modes prevents the turnover and decay of $A_{\ell m}$ as $\Omega \rightarrow \Omega_\mathrm{ISCO}$.

Finally, we consider the relative power between the $\ell = m = 2$ and $\ell = m = 15$ modes for a given orbital frequency. In the bottom panel of Figure \ref{fig:amplitudeVariation}, we plot the $A_{\ell, m = \ell}$ amplitudes as functions of $r_0$ for different values of $\hat{a}$. We see that $|A_{15,15}/A_{2,2}|^2 \lesssim 5\times 10^{-4}$ for all orbital frequencies and spins, placing a conservative upper bound for the relative power contributed by the $\ell = m = 15$ mode at a fixed frequency. For most year-long EMRI signals, the relative \emph{total} power (i.e., the power integrated across the entire frequency evolution of an inspiral) contributed by the $\ell = m = 15$ mode is $\lesssim 10^{-6}$, as we demonstrate in the following section. Therefore, we find it unnecessary to include higher harmonic contributions beyond $\ell \leq 15$.

\begin{figure}[!tp]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/amp_combined.pdf}
    \caption{Waveform amplitudes $A_{\ell, m}$ as a function of the radius of the orbit $r_0/M$. The top panel plots $A_{\ell=2, m=2}$ for different Kerr spin parameters $\hat{a}\geq 0.9$. The middle and bottom panels plot $A_{\ell, m=2}$ and $A_{\ell, m=\ell}$, respectively, for $\hat{a} = 0.9999$ and $2 \leq \ell \leq 15$. In the two lower panels, the mode magnitudes are ordered inversely with $\ell$. In other words, the $\ell = 2$ mode is the dominant mode, the $\ell = 3$ is the next most dominant mode, then $\ell =4$, $\ell = 5$, and so on, with $\ell = 15$ being the mode with the smallest magnitude (and darkest solid line). }
    %The bottom (yellow) line in the top panel is identical to the top (yellow) lines in the lower two panels.
    \label{fig:amplitudeVariation}
\end{figure}

@@

We presented the theoretical and numerical methods behind \texttt{bhpwave}: a new Python-based adiabatic gravitational waveform generator for binary sources composed of a compact object undergoing a quasi-circular inspiral into a rotating massive black hole. To build this waveform model, we precomputed mass-independent trajectories for systems with initial separations $r_0/M \lesssim 60$ and Kerr spins $|a| \leq 0.9999$. Furthermore, we precomputed all potential waveform harmonic amplitudes for $\ell \leq 15$. By implementing the $E(3)$ boundary conditions in our numerical spline algorithm, we improved the precision of our interpolated flux data and trajectories by as much as three orders of magnitude over traditional spline methods. We compute waveforms in both the time and frequency domains, and we find good agreement between these models. Furthermore, we compared our model against the FEW waveform generator for $\hat{a} = e_0 = 0$ and achieved mismatches $\mathcal{M} \sim 10^{-5}$. Using Fisher matrix calculations, we also assessed the magnitudes of any systematic biases introduced by numerical error in our model. We found that biases due to interpolation error are well below the thresholds required for LISA data analysis. Additionally, we demonstrated that waveform dephasing does not provide a complete picture of modeling error. In particular, for systems with retrograde orbits and slowly-rotating massive black holes, we found that waveform models could have phase errors of up to 10 radians, and yet the biases introduced by these errors would not be measurable by LISA. Thus, the claim that waveform models require subradian phase accuracy for LISA is a useful guide for model fidelity, but more sophisticated analyses are required to understand the true impact of modeling errors on LISA science (see also \cite{BurkETC23}).

A notable limitation of \texttt{bhpwave} is that it currently neglects the effects of eccentricity and precession. Nonetheless, while most observed EMRIs are expected to be highly eccentric \cite{GairETC04},  there are possible formation channels driven by accretion flow that circularize EMRI dynamics and align the orbital angular momentum with the massive black hole's spin \cite{PanLyuYang21}. Thus, \texttt{bhpwave} is applicable to these so-called ``wet-formation" EMRIs.
